State of the Practice of Lattice Boltzmann Method Software Development

### 2021-06-03 StateOfPractice/Peter-Notes/Report/Report.tex
3rd edits to results section of LBM Report

Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web page. Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web page or within formal documentation. Maintainability and correctness of these instructions could be improved if they were limited to one location.

@@

Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory. This artifact is often found within a user manual, on a web page, or is mentioned in related publications. In the latter case the user may need to spend significant time to find this information.

@@

Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model. In total 18 software packages have a tutorial and 13 have a user manual. The tutorials vary in scope and substance, and 8 include an expected output to compare user results to. Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page. Some packages do not have a user manual but do have useful documentation distributed on their web-pages. Expected user characteristics are documented in 4 software packages. Users are typically scientists or engineers trying to test a new application. Their background is often physics, chemistry, biophysics, or mathematics. All but one of the packages have a user support model, and several of them have multiple avenues of user support. The most popular avenue of support is git, followed by email and forums. One software package has a FAQ page.

@@

Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP. Software packages with a high score have a tutorial and user manual, and sometimes have documented user characteristics, and have at least one user support model, with many having several.

Interviews with developers revealed several usability issues. Some users have misunderstood boundaries of LBM and computational fluid dynamics in general, and have combined or applied methods that are not physically sound together. Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems. The users do not realize the limitations of the methods, of the software, and of how much time and effort is required to properly model a problem that can technically be modeled by the software. The software is not as easy as out of the box plug and play. Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.

Some additional infrequent software usability issues were commented on. Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models. These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirement specifications.

@@

Software packages were reviewed for the presence of artifacts. Every type of artifact or file that is not a code file was recorded. The software packages were also reviewed for version numbers. This information could be used to better troubleshoot software issues and organize documentation. All but 3 software packages have source code release and documentation version numbers.

@@

Issue tracking is used in 22 software packages, 15 of which use git, 6 use email, and one uses SourceForge. Most software packages that use git have most of their issues closed, and only 3 have less than 50 percent of their issues closed. Furthermore, most packages that use git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and 8 packages do not appear to use any such system.

Software package code files were further measured for the percentage of code that is comments. Packages with a higher percentage of comments were designated as more maintainable. Comments represent more than 10 percent of code files in 15 packages, and the average percentage of comments is 13.7.

@@

Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP. Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system. These packages also appear to reasonably handle tracked issues, having most of their issues closed. Their code files are well commented with more than 10 percent of the code being comments.

Interviews with developers revealed that most projects do have a defined process for accepting contributions from current team members or those that at some time had been linked to the packages or developers. The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.

@@

Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future. They would be very happy to see spontaneous contribution from outside of their organization, but now it is currently not the case.

Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction. Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported and have preferred dependencies. Data structures have also been redeveloped and storage has been improved. One software developer mentioned that the mathematical foundations that the geometries and models are designed from have been re-evaluated and optimized.

@@

A high degree of code modularity and abstraction was noted by one developer as a measure to ensure ease of future changes. They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package. Another developer noted that their software package was designed to easily allow for the addition of some LBM features, but changes to major aspects of the LBM models would be difficult. Furthermore, the package was designed with flexible data structures and storage in mind.

Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions. The Palabos package identifies as as a development framework for modeling problems in various computational fluid dynamics areas.

@@

The total number of source code files was measured for each software package. A larger number of source files was associated with increased reusability since this generally represented either more software package functionality or better modularization. The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.

@@

Figure \ref{Fig_Reusabilty} shows the reusability ranking of the software packages using AHP. Software packages with a high score have thousands of source code files and API documentation. The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM. For this reason a comparison between other software packages is not on a level field, but regardless of this these packages do have an abundance of reusable components.

There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}. The software design has enabled for waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.

@@

Ten random source code files of each software package were reviewed for several measures. This measure of surface understandability may not perfectly reflect each package due to this limitation. All of the packages appear to have a consistent indentation and formatting style. Only LUMA and HemeLB explicitly identify coding standards that are used during development. Generally, the software packages use consistent, distinctive, and meaningful code identifiers. Only 4 packages appear to use vague identifiers. Hard coded constants were observed in the source code of 12 packages. The constants are used for various parameters, mathematical constants, and matrix definitions. All of the packages are well commented. Comments clearly indicate what is being done. Domain algorithms are noted in the source code of 11 packages. All of the packages were well modularized. When observing the source code files, it was found that 13 of the packages had a consistent style and order for function parameters.

@@

Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP. Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive and meaningful code identifiers. They also have hard coded constants, and explicitly identify mathematical and LBM algorithms. Their comments are clear and indicative of what is being done. The source code is well modularized and structured.

Software developers noted that in their opinion users have generally found their packages to be understandable. One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as update the documentation to prevent misunderstanding the methods. Similarly, another developer noted that some users had issues setting up parameters for LBM schemes. The solution was to this was to update the interface where these parameters are set, as well as adding functionality to test the stability of the parameters. A third developer noted that some users lack the background knowledge to easily model flow problems using their software. A frequently asked questions section was added to the source package website to help users find answers to common questions. The package also has detailed documentation, including guides and usage requirements specification.

@@

These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}. The following type of development tools were explicitly noted in the artifacts or web pages of the 23 LBM packages that were assessed. It is likely that other tools, such as debuggers, were used but are not specified here.

@@

The following types of dependencies were explicitly noted in the artifacts or web pages of the 23 LBM packages that were assessed. It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.

@@

Many of the software packages that were assessed were developed by teams of two or more people. Their work needed to be coordinated and managed. The following types of project management tools were explicitly noted in the artifacts, web pages, or interviews with the developers of the 23 LBM packages that were assessed. As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.

@@

\def\arraystretch{1.5}
\begin{tabular}{p{16cm}}
		\hline		
		\textbf{Installability  (Measured via installation on a virtual machine.) }\\
		\hline
		Are there installation instructions? ({yes, no})\\
		Are the installation instructions in one place? Place referring to a single document or web page. ({yes, no, n/a})\\
		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
		Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
		Are compatible operating system versions listed? ({yes, no})\\
		Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
		If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
		Is there a specified way to validate the installation? ({yes*, no})\\
		How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
		What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
		How many extra software packages need to be installed before or during installation? (number)\\
		Are required package versions listed? ({yes, no, n/a})\\
		Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
		Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
		Overall impression? ({1 .. 10})\\
		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
		\hline
\end{tabular}

@@

\captionof{table}{Grading Template}\label{gradingtemplate}
\def\arraystretch{1.5}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
	\hline
	Are there installation instructions? ({yes=1, no=-1})\\
	Are the installation instructions in one place? Place referring to a single document or web page. ({yes=1, no=0, n/a=0})\\
	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
	Are compatible operating system versions listed? ({yes=1, no=0})\\
	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
	What OS was used for the installation? (does not count)\\
	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
	Overall impression? (a sum of $>$10 is rounded down to 10)\\
	\hline
\end{tabular}

>>>>

Of the software packages that were tested, most installation instructions are located in one place, often in an instruction manual or on a web-page. Sometimes incomplete installation instructions are found on a home page, with more detailed instructions located on another web-page or within formal documentation. Maintainability and correctness of these instructions could be improved if they were limited to one location.

@@

Sixteen of the software packages include a requirements specification artifact or explicitly reference domain theory. This artifact is often found within a user manual, on a web-page, or is mentioned in related publications. In the latter case the user may need to spend significant time to find this information.

@@

Software package artifacts were reviewed for the presence of a tutorial, a user manual, documented user characteristics, and a user support model. In total eighteen software packages have a tutorial and thirteen have a user manual. The tutorials vary in scope and substance, and eight include an expected output. Most user manuals are in the form of a file that can be downloaded, while some are rendered on a web-page. Some packages do not have a user manual but do have useful documentation distributed on their web-pages. Expected user characteristics are documented in four software packages. Users are typically scientists or engineers trying to test a new application. Their background is often physics, chemistry, biophysics, or mathematics. All but one of the packages have a user support model, and several of them have multiple avenues of user support. The most popular avenue of support is Git, followed by email and forums. One software package has a FAQ page.

@@

Figure \ref{Fig_Usability} shows the surface usability ranking of the software packages using AHP. Software packages with a high score have a tutorial and user manual, sometimes have documented user characteristics, and have at least one user support model. Many packages have several user models.

Interviews with developers revealed several usability issues. Some users have misunderstood the boundaries of LBM and computational fluid dynamics, and have combined or applied methods that are not physically sound. Sometimes users have applied LBM to poorly defined or inappropriate fluid dynamics problems. The users do not realize the limitations of the methods, of the software, and do not understand the requirements to properly model a problem with the software. The software packages are not designed to be used out of the box in a plug and play fashion. Developers of some of the software packages mitigated this by editing the source code to prevent unsound combination from being applied, and by updating the documentation to better inform users of LBM limitations, and of the requirements to properly model appropriate problems, including what algorithms and parameters to use.

Some additional but infrequent software usability issues were commented on by the developers. Users have had trouble with installation and understanding how to maneuver the interfaces and how to set up or run models. These issues are addressed by various user support models, including frequently asked questions sections on the software package websites, user guides, and hardware and software requirements specifications.

@@

Software packages were reviewed for the presence of artifacts. Every type of artifact or file that is not a code file was recorded. The software packages were also reviewed for software release and documentation version numbers. This information could be used to better troubleshoot issues and organize documentation. All but three software packages have source code release and documentation version numbers.

@@

Issue tracking is used in twenty two software packages, fifteen of which use Git, six use email, and one uses SourceForge. Most software packages that use Git have most of their issues closed, and only three have less than fifty percent of their issues closed. Furthermore, most packages that use Git for issue tracking use GitHub as a version control system, one package uses CVS for issue tracking, and eight packages do not appear to use any such system.

Software package code files were further measured for the percentage of code that is comments. Packages with a higher percentage of comments were designated as more maintainable. Comments represent more than ten percent of code files in fifteen packages, and the average percentage of code comments is about fourteen.

@@

Figure \ref{Fig_Maintainability} shows the maintainability ranking of the software packages using AHP. Software packages with a high score provide version numbers on documents and source code releases, have an abundance of high quality artifacts, and use an issue tracking tool and version control system. These packages also appear to reasonably handle tracked issues, having most of their issues closed. Their code files are well commented with more than ten percent of the code being comments.

Interviews with developers revealed that most projects do have a defined process for accepting contributions from team members. The packages rarely get contributions from outside developers, but the process would be similar as for the aforementioned group.

@@

Some developers noted that their software package does not have well defined contributing guide in the repository but it might be a good idea to add one in the near future. They would be very happy to see contribution from outside of their organization, but currently this does not happen.

Furthermore, maintainability has been addressed by increasing source code modularity, reducing duplicate information, and improving abstraction. Several software packages have had large sections of their code base redeveloped with languages that the developers felt are more understandable and readable, and that are better supported. Data structures have also been redeveloped and storage has been improved. One software developer mentioned that the mathematical foundations that the geometries and models are designed from had been re-evaluated and optimized.

@@

A high degree of code modularity and abstraction was noted by one developer as a measure to ensure the ease of future changes. They also noted that some of the code base was transitioned from c to c++, which could ease modifiability of that software package. Another developer noted that their software package was designed to easily allow for the addition of some LBM features, but changes to major aspects of the LBM models would be difficult. Furthermore, the package was designed with flexible data structures and storage in mind.

Some software packages, like Palabos, provide validation benchmarks for their core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated intent of these benchmarks is to showcase the validity and usefulness of the package in order to stimulate the development of third-user extensions. The Palabos package identifies as a development framework for modeling problems in various computational fluid dynamics areas.

@@

Each software package was measured for the total number of source code files. A larger number of source files was associated with increased reusability due to increased modularization. The software packages were also reviewed for the presence of API documentation, which indicates that a software package was developed with interaction between other software applications in mind.

@@

Figure \ref{Fig_Reusabilty} shows the reusability ranking of the software packages using AHP. Software packages with a high score have thousands of source code files and API documentation. The highest scoring packages, ESPResSo and waLBerla, have extensive functionality, including graphical visualizations as well as modeling that does not use LBM. For this reason a comparison with other software packages is not on a level field. However, these packages do have an abundance of reusable components.

There was a strong focus on modularity when designing the waLBerla framework in order to enhance productivity, reusability, and maintainability \citep{bauer2021walberla}. Its software design has enabled waLBerla to be successfully applied in several projects as a basis for various extensions \citep{bauer2021walberla}.

@@

Ten random source code files of each software package were reviewed for several measures. This measure of surface understandability may not perfectly reflect each package due to the limitation of only testing ten files.

All of the packages appear to have a consistent indentation and formatting style. Only LUMA and HemeLB explicitly identify coding standards that are used during development. Generally, the software packages use consistent, distinctive, and meaningful code identifiers. Only four packages appear to use vague identifiers. Hard coded constants were observed in the source code of twelve packages. The constants are used for various parameters, mathematical constants, and matrix definitions. All of the packages are well commented and the comments clearly indicate what is being done. Domain algorithms are noted in the source code of eleven packages. All of the packages are well modularized. When observing the source code files, it was found that thirteen of the packages have a consistent style and order for function parameters.

@@

Figure \ref{Fig_Understandability} shows the surface understandability ranking of the software packages using AHP. Software packages with a high score have a consistent indentation and formatting style, and consistent, distinctive, and meaningful code identifiers. They also have hard coded constants, and explicitly identify mathematical and LBM algorithms. Their comments are clear and indicate what is being done in the source code. The source code is well modularized and structured.

Software developers noted that in their opinion users have generally found their packages to be understandable. One developer commented that some users have attempted to run physically incompatible LBM methods, and the solution was to edit the code to prevent such combinations, as well as to update the documentation to prevent misunderstanding the methods. Similarly, another developer noted that some users had issues setting up parameters for LBM schemes. The solution to this was to update the interface where these parameters are set, as well as to add functionality to test the stability of the parameters. A third developer noted that some users lack the background knowledge to easily model fluid dynamics problems using their software. A frequently asked questions section was added to their package website to help users find answers to common questions. The package also has detailed documentation, including guides and usage requirements specification, to better help user understand the software.

@@

These tools support the development of end products but do not become part of them, unlike dependencies that remain in the application once it is released \citep{ghezzi1991fundamentals}. The following type of development tools were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed. It is likely that other tools, such as debuggers, were used but are not specified here.

@@

The following types of dependencies were explicitly noted in the artifacts or web-pages of the 23 LBM packages that were assessed. It is possible that other types of dependencies are part of these software packages but are not clearly specified in their artifacts or web sites and because of that they are not listed here.

@@

Many of the software packages that were assessed were developed by teams of two or more people. Their work needed to be coordinated and managed. The following types of project management tools were explicitly noted in the artifacts, web-pages, or interviews with the developers of the 23 LBM packages that were assessed. As with development tools and dependencies, it is possible that other types of project management tools were used to coordinate and manage the projects but are not specified and because of that they are not listed here.

@@

\def\arraystretch{1.5}
\begin{tabular}{p{16cm}}
		\hline		
		\textbf{Installability  (Measured via installation on a virtual machine.) }\\
		\hline
		Are there installation instructions? ({yes, no})\\
		Are the installation instructions in one place? Place referring to a single document or web-page. ({yes, no, n/a})\\
		Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes, no, n/a})\\
		Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes, no, unclear})\\
		Are compatible operating system versions listed? ({yes, no})\\
		Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*, no})\\
		If the software installation broke, was a descriptive error message displayed? ({yes, no, n/a})\\
		Is there a specified way to validate the installation? ({yes*, no})\\
		How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. (number, OS)\\
		What OS was used for the installation? ({Windows, Linux, OS X, Android, other* }) *given via string\\
		How many extra software packages need to be installed before or during installation? (number)\\
		Are required package versions listed? ({yes, no, n/a})\\
		Are there instructions for the installation of required packages / dependencies? ({yes, no, n/a})\\
		Run uninstall, if available. Were any obvious problems caused? ({yes∗ , no, unavail})\\
		Overall impression? ({1 .. 10})\\
		Additional comments? (can cover any metrics you feel are missing, or any other thoughts you have)\\
		\hline
\end{tabular}

@@

\captionof{table}{Grading Template}\label{gradingtemplate}
\def\arraystretch{1.5}
\begin{tabular}{p{16cm}}
	\hline		
	\textbf{Installability  (Measured via installation on a virtual machine.) }\\
	\hline
	Are there installation instructions? ({yes=1, no=-1})\\
	Are the installation instructions in one place? Place referring to a single document or web-page. ({yes=1, no=0, n/a=0})\\
	Are the installation instructions linear? Linear meaning progressing  in a single series of steps. ({yes=1, no=0, n/a=0})\\
	Are the instructions written as if the person doing the installation has none of the dependent packages installed? ({yes=1, no=0, unclear=0})\\
	Are compatible operating system versions listed? ({yes=1, no=0})\\
	Is there something in place to automate the installation (makefile, script, installer, etc)? ({yes*=1, no=-1})\\
	If the software installation broke, was a descriptive error message displayed? ({yes=0, no=-2, n/a=1})\\
	Is there a specified way to validate the installation? ({yes*=1, no=0})\\
	How many steps were involved in the installation? (Includes manual steps like unzipping files) Specify OS. ($<$10 = 1)\\
	What OS was used for the installation? (does not count)\\
	How many extra software packages need to be installed before or during installation? ($<$10 = 1)\\
	Are required package versions listed? ({yes=1, no=0, n/a=1})\\
	Are there instructions for the installation of required packages / dependencies? ({yes=1, no=0, n/a=1})\\
	Run uninstall, if available. Were any obvious problems caused? ({yes∗=0, no=1, unavail=1})\\
	Overall impression? (a sum of $>$10 is rounded down to 10)\\
	\hline
\end{tabular}
